Analysis-ready VCF at Biobank scale using Zarr

### 2024-05-30 paper.tex

Fig~\ref{fig-compression-chunksize} shows that chunk size has a 
weak influence on compression ratio for most fields. Interestingly,
the choice of chunk size has a somewhat unpredictable effect
on compression for the genotype matrix, suggesting that how 
samples are grouped together has a strong effect on compression.
At around 50X, compression is excellent in all cases however, 
and given that genotypes contribute relatively little to the
overall storage cost in many datasets 
(Table~\ref{tab-genomics-england-data}), the differences are slight.
Otherwise, increasing sample chunk size slightly increases 
compression on call\_AB, and has no effect on less compressible
fields. Variant chunk size appears to have almost no effect on compression 
ratio. Thus, we do not expect the choice of chunk size to have a significant
impact on overall storage usage, and so choice should be
determined by other considerations such as expected data
access patterns.

@@

Fig~\ref{fig-compression-compressor} shows the effect of varying compression
codecs in Blosc. The combination of outstanding compression performance
and good decoding speed makes zstd a good default choice.

@@

\begin{figure}
\includegraphics{figures/subset-matrix-compute-supplemental}
\caption{Compute performance on a large subset of the genotype matrix.
Total CPU time required to run the af-dist calculation for
a subset of half of the samples and 10000 variants from the middle of the matrix
for the data in Figure~\ref{fig-data-storage}.
Genozip did not run for
$n > 10^4$ samples because it does not support a file to specify
sample IDs, and the command line was therefore too long for the shell
to execute. 
\label{fig-subset-matrix-compute-supplemental}}
\end{figure}

@@

\begin{figure}
\includegraphics{figures/compression-chunksize-finegrained.pdf}
\caption{Effects of sample chunk size on compression ratio on call\_genotype
field in 1000 Genomes data.
The same analysis as in Fig~\ref{fig-compression-chunksize}, except we only
consider call\_genotype and we consider all sample chunk sizes from 
100 to 256. Distinct trend-lines emerge for odd, even and multiple-of-four
chunk sizes (shown by markers). The size of the final chunk also has a minor
effect (shown by colour).
\label{fig-compression-chunksize-finegrained}}
\end{figure}

\begin{figure}
\includegraphics{figures/compression-compressor}
\caption{Effects of Blosc compression codec on compression ratio on call-level 
fields in 1000 Genomes data. 
In all cases compression level=7 was used, with a variant
chunk size of 10,000 and sample chunk size of 1,000.
Bit shuffle was used for call\_genotype, and no shuffle used for the other fields.
\label{fig-compression-compressor}}
\end{figure}

>>>>

Fig~\ref{fig-compression-compressor} shows the effect of varying compression
codecs in Blosc. The combination of outstanding compression performance
and competitive decoding speed (Fig~\ref{fig-whole-matrix-decode}) 
makes zstd a good default choice.

@@

Fig~\ref{fig-compression-chunksize} shows that chunk size has a 
weak influence on compression ratio for most fields. 
Increasing sample chunk size slightly increases 
compression on call\_AB, and has no effect on less compressible
fields. Variant chunk size appears to have almost no effect 
on compression ratio. 
Interestingly, the choice chunk size along the sample dimension for the
genotype matrix does a significant effect. 
With six evenly spaced points between 100 and 2504,
Fig~\ref{fig-compression-chunksize}A shows a somewhat unpredictable 
relationship between sample chunk size and compression ratio.
The more fine-grained analysis of
Fig~\ref{fig-compression-chunksize-finegrained} shows that three
distinct trend lines emerge depending on the chunk size 
divisibility, with the modulus 
(i.e., the remainder in the last chunk) also having a minor effect. 
At greater than 40X,
compression ratio is 
high in all cases, and given that genotypes contribute relatively
little to the total storage of real datasets
(Table~\ref{tab-genomics-england-data}) the effect will
likely be fairly minor in practice.
Thus, we do not expect the choice of chunk size to have a significant
impact on overall storage usage, and so choice may be
determined by other considerations such as expected data
access patterns.

@@

\begin{figure}
\includegraphics{figures/subset-matrix-compute-supplemental}
\caption{Compute performance on a large subset of the genotype matrix.
Total CPU time required to run the af-dist calculation for
a subset of half of the samples and 10000 variants from the middle of the matrix
for the data in Figure~\ref{fig-data-storage}.
Genozip did not run for
$n > 10^4$ samples because it does not support a file to specify
sample IDs, and the command line was therefore too long for the shell
to execute. 
\label{fig-subset-matrix-compute-supplemental}}
\end{figure}
\begin{figure}

\includegraphics{figures/compression-compressor}
\caption{Effects of Blosc compression codec on compression ratio on call-level 
fields in 1000 Genomes data. 
In all cases compression level=7 was used, with a variant
chunk size of 10,000 and sample chunk size of 1,000.
Bit shuffle was used for call\_genotype, and no shuffle used for the other fields.
\label{fig-compression-compressor}}
\end{figure}

@@

\begin{figure}
\includegraphics{figures/compression-chunksize-finegrained.pdf}
\caption{Effects of sample chunk size on compression ratio on the call\_genotype
field in 1000 Genomes data.
The same analysis as in Fig~\ref{fig-compression-chunksize}, except we only
consider call\_genotype and we examine all sample chunk sizes from 
100 to 256. Distinct trend-lines emerge for odd, even and multiple-of-four
chunk sizes (shown by markers). The size of the final chunk also has a minor
effect (shown by colour).
\label{fig-compression-chunksize-finegrained}}
\end{figure}
